The interactions of rational, pragmatic agents\\ lead to efficient language structure and use

### 2019-01-31 paper/paper.tex
updates from MF / BP edits

\begin{abstract}
Languages display a diverse set of distributional regularities such as
the relation between a word's frequency and rank in a corpus, the
distribution of syntactic dependency lengths across languages, or the
presence of properties such as ambiguity. We discuss a framework for
studying how these system-level properties emerge from local,
in-the-moment interactions of rational, pragmatic speakers and
listeners. To do so, we derive a novel objective function for measuring
the communicative efficiency of linguistic systems in terms of the
interactions of speakers and listeners. We examine the behavior of this
objective in a series of simulations focusing on the communicative
function of ambiguity in language. These simulations suggest that
rational pragmatic agents will produce communicatively efficient systems
and that interactions between such agents provide a framework for
examining efficient properties of language structure and use more
broadly.

@@

But how does this optimization take place? The example given by Zipf
(1949), describes local, communicative interactions in terms of a
\textit{reference game}. Speakers intend to refer to some object in the
world \(m\). They choose some utterance \(u\) to transmit this intended
meaning, \(u \rightarrow m\). The listener attempts to reconstruct this
intended meaning given the transmitted utterance, \(m \rightarrow u\).
Other projects have assumed this basic reference game setting
(Piantadosi et al., 2011; Regier et al., 2015) and this simplification
of the communicative act has proven productive in theoretical
(Ferrer-i-Cancho, 2018), simulation-based (Kirby et al., 2014) and
empirical explorations (Hawkins, Franke, Smith, \& Goodman, 2018) of
efficient language structure and use.\par

Zipf's conception of speaker and listener effort may be connected to
speakers' pragmatic reasoning in the moment. Under a Gricean treatment
of pragmatics, speakers and listeners follow a set of conversational
maxims in which they cooperate to transfer information (Grice, 1975).
These maxims appear to emerge from efficiency concerns, however (Horn,
1984). We formalize this connection -- showing how system-level
efficiencies can emerge from local interaction behavior. Intuitively,
our claim is that to understand an ``efficient'' property of a system it
is essential that we consider how that property is \emph{used}
efficiently. It follows that any quantitative measure of ``efficiency''
must be a function of both the system and its users.\par

In the current work we provide a case-study for the framing presented
above. We choose a property of languages that could, in principle, vary
freely, but shows strong regularities across languages. The explanadum
is why this regularity holds. We examine ambiguity as our property,
extending ideas by Piantadosi et al. (2011). We define a novel measure
of efficiency that depends on the interactional behavior of speaker and
listener agents. We adopt the reference game as our primary unit of
interaction and model language users with the Rational Speech Act (RSA)
framework -- a computional model of language use, which is nicely
attested by experimental data on interaction. Using these basic
ingredients, we show that the property of interest (ambiguity) is
prevalent in languages that optimize our measure of efficiency. Further,
we show how ambiguity is \textit{used} efficiently during local,
in-the-moment interactions. The contributions of this work are twofold
-- we derive a novel objective for measuring linguistic efficiency and
also show how the reference game framework, in combination with formal
models of communication, can be used to connect ideas about system-level
efficiencies to in-the-moment language use.\par

@@

The Rational Speech Act framework (RSA) is a computational-level theory
of pragmatic language use, which has been well attested in experimental
work examining both speaker and listener communication behavior (Frank
\& Goodman, 2012; Goodman \& Frank, 2016). RSA is a formalization of
essential Gricean pragmatic principles -- agents reason about one
another and their shared context (Grice, 1975). We adopt RSA as our
representational framework to model Gricean (rational and pragmatic)
speaker-listeners in the reference game
setting\footnote{See SI section 1 for a more detailed description of basic RSA speaker-listener defintions and to Goodman \& Frank (2016) for an overview of its applications.}.\par

@@

From an information-theoretic perspective this objective is intuitive:
The Cross-Entropy (CE) gives us a measure of dissimilarity between two
distributions -- the average number of bits required to communicate
under one distribution, given that the ``true'' distribution differs. In
our case, we have an expectation over this term -- the expected
difference between the distributions assumed by the speaker
\(P_{speaker}\) and listener \(P_{listener}\) given a set of contexts
\(C\)\footnote{Note that in the single context case $|C|=1$ this objective is simply the speaker-listener Cross-Entropy over utterance-meaning pairs $H_{cross}[P_{speaker}(u, m; \ell), P_{listener}(u, m; \ell)]$}.
In other words, an ``efficient'' language \(\ell\) minimizes the
distance between what speakers and listeners think.

@@

In the following experiments we explore two aspects of Piantadosi's
claim. In Simulation 1, we examine the efficient language
\emph{structure} aspect of their claim, exploring when the optimal
linguistic system \(\ell^*\) is most likely to contain ambiguous lexical
items. In Simulation 2, we explore an efficient language \emph{use}
aspect of the claim -- at what point in a conversation is it useful for
a speaker to use ambiguous lexical material?\par

@@

We adopt the hypothesis presented by Piantadosi, formalizing their
argument in our current framework -- showing that ambiguity is an
efficient property under our CE objective in the reference game setting.
We proceed by generating languages with different amounts of contextual
support (varying the size of \(|C|\)). We search the space of languages,
examining whether ones which minimize our objective contain ambiguity.
If context leads to more efficient communication, then optimal languages
under our objective should be more likley to be ambiguous as the amount
of context increases.\par

@@

Following the argument given by Piantadosi et al. (2011), we want to
assess the impact of \emph{context} on the presence of ambiguity in
optimal languages. To do so we consider four conditions with \(n=500\)
simulations each (that is, 500 unique sets of
\(\{P(U), P(M|C), P(C)\}\). Our first is a \textit{one-context}
condition (\(|C|=1\)) -- only a single distribution over meanings
\(P(M)\). In our \textit{two-context} condition (\(|C| = 2\)), we
consider efficiency under both \(P(M|C=c_1)\) as well as \(P(M|C=c_2)\).
\textit{Three-} and \textit{four-context} conditions corresponding
accordingly.\par

In Simulation 1 we explored the degree to which ambiguity is an
efficient property of languages when communication is contextualized.
Figure 1. plots the proportion of optimal languages under each objective
as a function of context size. The red line shows that as the number of
contexts increases, so does the probability that an optimal language
\(\ell^*_{cross}\) contains ambiguity under our CE objective. For
comparison we also plot the proportion of speaker-optimal
\(\ell^*_{speaker}\) (blue line) and listener-optimal
\(\ell^*_{listener}\) (green line) languages that contain ambiguity. In
line with Zipf's predictions, if languages are designed only to minimize
speaker effort then optimal languages always contain ambiguity. If
languages are designed to minimize listener effort then ambiguity is
always avoided.\par

While our results indicate that ambiguity is an efficient property of
contextualized language, these simulations assumed that agents could
always disambiguate items in context. That is, in our four conditions
both agents had perfect knowledge of the relevant conditional
distributions (\(P(M|C)\)). This assumption may be too strong for
describing much of day-to-day communication -- we seldom interact with
others with perfect knowledge of the current context (or topic) at the
start of a conversation. To explore how ambiguity may be \textit{used}
efficiently in our framework, we next examine a case in which the
listener has imperfect knowledge of context at the start of the
conversation, but may infer it from the contents of discourse.\par

We shift focus in Simulation 2 -- instead of examining the landscape of
a space of languages under our CE objective, we consider a single
language containing ambiguity, examining how that property is used over
a discourse. If context is not informative with respect to meaning early
in a discourse, but is at later positions, then an efficient speaking
strategy would avoid ambiguous material early, saving it for when
contextual information is disambiguating. We can consider ``context'' as
analogous to a ``topic'' of conversation. As an example, imagine a
scenario in which a reader is beginning a magazine article. While they
may have some knowledge about the article's topic (perhaps from the
title), they may not have complete knowledge of its contents, including
the persons or events involved. In this setting, using a low-cost, but
ambiguous pronoun early may lead to misunderstanding if context is not
informative. But, if by a later position enough contextual information
has accumulated, it may be efficient to use the ambiguous material.\par

We consider consider a single language \(\ell\), which contains both
ambiguous and unambiguous utterances. We assume ambiguous utterances are
lower cost. Crucially, we do not assume that the listener knows the
particular topic (\(c_{current}\)) of the conversation
\textit{a priori}. Rather, that the listener has knowledge of the set of
possible topics \(C = \{c_1, \dots, c_k\}\), but
\textit{does not know which one is currently being used by the speaker}.
Formally, this means the listener does not have access to the correct
conditional distribution over meanings \(P(M|C=c_{current})\) at the
start of the discourse.\par

@@

We conduct \(N=600\) simulations, generating discourses of length
\(|D|=30\) utterances, comparing three speaker models (\(n=200\) each).
We consider a single language
\(\ell\)\footnote{See SI for the matrix notation of this langauge.} with
\(|U|=6\) and \(|M|=4\) in which two of the utterances are ambiguous and
lower cost than the unambiguous utterances. (Note that use of this
particular language is not essential -- the results are broadly
generalizable to languages that contain ambiguity.)\par

@@

We consider three types of speaker models. Our \textit{Full pragmatics}
agent, models a speaker who reasons about her listener and also has
complete recall of the set of utterances in the discourse \(D\). This
speaker believes that the listener may not know the current topic
\(c_{current}\) at the start of the discrouse, but can infer it over the
discourse. We compare two baseline models. The first, a
\textit{Partial pragmatics} baseline describes a speaker who reasons
about a listener, but assumes they have no access to discourse history.
The second, a \textit{No pragmatics} speaker who does not consider a
listener at all, but produces utterances according to the underlying
langauge semantics (\(\ell\)) and topic probabilities
(\(p(M|C=c_{current}\)) (see SI section 3 for details).\par

@@

Figure 2, (A) shows the empirical probability that a speaker uses an
ambiguous utterance as a function of discourse position. The
\textit{No pragmatic} baseline uses ambiguous utterances frequently and
at a constant rate over the discourse and the \textit{Partial pragmatic}
baseline avoids ambiguous utterances entirely. But, the
\textit{Full pragmatic} model avoids ambiguous material only at the
start of the discourse, employing it increasingly as the discourse
proceeds. (C) tracks our CE objective for each model over the discourse.
Note that the objective decreases for all three models, primarily driven
by the listener updating his belief about the actual topic
(\(P(C=c_{current}|D)\)). However, the objective declines more quickly
under the \textit{Full} and \textit{Partial} pragmatic speakers as
listener agents are better able to infer the correct context.
Additionally, the difference in CE between the \textit{Full}- and
\textit{Partial pragmatic} models at the end of the discourse is driven
by the reduction in speaker costs described in (B). While speaker effort
remains constant in both \textit{No pragmatic} and
\textit{Partial pragmatic} baselines, effort declines in the
\textit{Full pragmatic} model as she increasingly relies on ambiguous
material later in the discourse.\par

@@

How do the competing pressures of speakers and listeners give rise to
the distributional forms found in natural language? Zipf (1949) proposed
that the asymmetry between speaker and listener costs gives rise to a
range of properties at the level of the lexicon. We explored the
interactions of rational pragmatic agents as a framework for
understanding efficient language structure and use. We focused on an
argument on the communicative function of ambiguity (Piantadosi et al.,
2011), deriving a novel speaker-listener Cross-Entropy objective for
measuring the efficiency linguistic systems from first principles of
efficient language use. In a series of simulations we showed that
optimal languages are more likely to contain ambiguous material when
context is informative and that rational pragmatic agents will
\emph{use} ambiguous material efficiently.\par

A limitation of the current work is an analysis of exactly how this
objective compares to existing measures. For example, previous work has
described competing speaker-listener pressures in terms of a trade-off
of \textit{simplicity} and \textit{informativeness} (Kemp \& Regier,
2012) or \textit{expressivity} and \textit{compressibility} (Smith,
Tamariz, \& Kirby, 2013). Future work should assess the degree to which
we can derive the same properties as previous studies using our current
framework. In line with this idea, a primary focus of the current work
is to provide a case-study in the productivity of assessing
functionalist theories in repeated reference games between rational
pragmatic agents. In theory, this framework can serve as a domain
general tool to assess the range of functionalist theories examining
efficient language-structure and use.\par

>>>>

\begin{abstract}
Despite their diversity, languages around the world share a consistent
set of properties and distributional regularities. For example, the
distribution of word frequencies, the distribution of syntactic
dependency lengths, and the presence of ambiguity are all remarkably
consistent across languages. We discuss a framework for studying how
these system-level properties emerge from local, in-the-moment
interactions of rational, pragmatic speakers and listeners. To do so, we
derive a novel objective function for measuring the communicative
efficiency of linguistic systems in terms of the interactions of
speakers and listeners. We examine the behavior of this objective in a
series of simulations focusing on the communicative function of
ambiguity in language. These simulations suggest that rational pragmatic
agents will produce communicatively efficient systems and that
interactions between such agents provide a framework for examining
efficient properties of language structure and use more broadly.

@@

But how does this optimization take place? The example given by Zipf
(1949) describes local, communicative interactions in terms of a
\textit{reference game}. Speakers intend to refer to some object in the
world \(m\). They choose some utterance \(u\) to transmit this intended
meaning, \(u \rightarrow m\). The listener attempts to reconstruct this
intended meaning given the transmitted utterance, \(m \rightarrow u\).
Other projects have assumed this basic reference game setting
(Piantadosi et al., 2011; Regier et al., 2015) and this simplification
of the communicative act has proven productive in theoretical
(Ferrer-i-Cancho, 2018), simulation-based (Kirby et al., 2014) and
empirical explorations (Hawkins, Franke, Smith, \& Goodman, 2018) of
efficient language structure and use.\par

Adopting reference games as a basic unit of analysis suggests that
optimization may take place at the level of conversation. Importantly,
Zipf's conception of speaker and listener effort may be connected to how
language is used; in particular, whether interlocutors engage in
pragmatic reasoning during conversation. Under a Gricean treatment of
pragmatics, speakers and listeners follow a set of conversational maxims
in which they cooperate to transfer information (Grice, 1975). These
maxims appear to emerge from efficiency concerns, however (Horn, 1984).
We formalize this connection -- showing how system-level efficiencies
can emerge from local interaction behavior of pragmatic agents.
Intuitively, our claim is that to understand an ``efficient'' property
of a system it is essential that we consider how that property is
\emph{used} efficiently.\par

We provide a case study for this approach, in which functionalist
regularities emerge from the dynamics of pragmatic communication. We
choose a property of languages that could, in principle, vary freely,
but shows strong regularities across languages. The explanadum is why
this regularity holds. We examine ambiguity as our property, extending
ideas by Piantadosi et al. (2011). We define a novel measure of
efficiency that depends on the interactional behavior of speaker and
listener agents. We adopt the reference game as our primary unit of
interaction and model language users with the Rational Speech Act (RSA)
framework -- a computional model of language use, which is nicely
attested by experimental data on interaction. Using these basic
ingredients, we show that the property of interest (ambiguity) is
prevalent in languages that optimize our measure of efficiency. Further,
we show how ambiguity is \textit{used} efficiently during local,
in-the-moment interactions. The contributions of this work are twofold
-- we derive a novel objective for measuring linguistic efficiency and
also show how the reference game framework, in combination with formal
models of communication, can be used to connect ideas about system-level
efficiencies to in-the-moment language use.\par

@@

The Rational Speech Act framework (RSA) is a computational-level theory
of pragmatic language use, which has produced good fit to human
communication behavior across a range of language phenomena (Frank \&
Goodman, 2012; Goodman \& Frank, 2016). RSA is a formalization of
essential Gricean pragmatic principles -- agents reason about one
another and their shared context (Grice, 1975). We adopt RSA as our
representational framework to model Gricean (rational and pragmatic)
speaker-listeners in the reference game setting (see SI and Goodman \&
Frank (2016) for review).\par

@@

From an information-theoretic perspective this objective is intuitive:
\(H_{cross}\) denotes the Cross-Entropy (CE), a measure of dissimilarity
between two distributions -- the average number of bits required to
communicate under one distribution, given that the ``true'' distribution
differs. In our case, we have an expectation over this term -- the
expected difference between the distributions assumed by the speaker
\(P_{speaker}\) and listener \(P_{listener}\) given a set of contexts
\(C\)\footnote{Note that in the single context case $|C|=1$ this objective is simply the speaker-listener Cross-Entropy over utterance-meaning pairs $H_{cross}[P_{speaker}(u, m; \ell), P_{listener}(u, m; \ell)]$}.
In other words, an ``efficient'' language \(\ell\) minimizes the
distance between what speakers and listeners think.

@@

In the following experiments we explore two aspects of Piantadosi's et
al.'s claim. In Simulation 1, we examine the efficient language
\emph{structure} aspect of their claim, exploring when the optimal
linguistic system \(\ell^*\) is most likely to contain ambiguous lexical
items. In Simulation 2, we explore an efficient language \emph{use}
aspect of the claim -- at what point in a conversation is it useful for
a speaker to use ambiguous lexical material?\par

@@

We show that ambiguity is an efficient property under our CE objective
in the reference game setting. We proceed by generating languages with
different amounts of contextual support (varying the size of \(|C|\)).
We search the space of languages, examining whether ones which minimize
our objective contain ambiguity. If context leads to more efficient
communication, then optimal languages under our objective should be more
likely to be ambiguous as the amount of context increases.\par

@@

@@

We want to assess the impact of \emph{context} on the presence of
ambiguity in optimal languages. To do so we consider four conditions
with \(n=500\) simulations each (that is, 500 unique sets of
\(\{P(U), P(M|C), P(C)\}\). Our first is a \textit{one-context}
condition (\(|C|=1\)) -- only a single distribution over meanings
\(P(M)\). In our \textit{two-context} condition (\(|C| = 2\)), we
consider efficiency under both \(P(M|C=c_1)\) as well as \(P(M|C=c_2)\).
\textit{Three-} and \textit{four-context} conditions corresponding
accordingly.\par

In Simulation 1 we explored the degree to which ambiguity is an
efficient property of languages when communication is contextualized.
Figure 1 plots the proportion of optimal languages under each objective
as a function of context size. The red line shows that as the number of
contexts increases, so does the probability that an optimal language
\(\ell^*_{cross}\) contains ambiguity under our CE objective. For
comparison we also plot the proportion of speaker-optimal
\(\ell^*_{speaker}\) (blue line) and listener-optimal
\(\ell^*_{listener}\) (green line) languages that contain ambiguity. In
line with Zipf's predictions, if languages are designed only to minimize
speaker effort then optimal languages always contain ambiguity. If
languages are designed to minimize listener effort then ambiguity is
always avoided.\par

While our results indicate that ambiguity is an efficient property of
contextualized language, these simulations assumed that agents could
always disambiguate items in context. That is, in our four conditions
both agents had perfect knowledge of the relevant conditional
distributions (\(P(M|C)\)). This assumption may be too strong for
describing much of day-to-day communication -- we seldom interact with
others with perfect knowledge of the current context (or topic) at the
start of a conversation. To explore how ambiguity may be \textit{used}
efficiently in our framework, we next examine a case in which the
listener has imperfect knowledge of context at the start of the
conversation, but may infer it from the discourse history.\par

We shift focus in Simulation 2 -- instead of examining the landscape of
a space of languages under our CE objective, we consider a single
language containing ambiguity, examining how that property is used over
a discourse. If context is not informative with respect to meaning early
in a discourse, but is at later positions, then an efficient speaking
strategy would avoid ambiguous material early, saving it for when
contextual information is disambiguating. We can consider ``context'' as
analogous to a ``topic'' of conversation. As an example, imagine a
scenario in which a reader is beginning a magazine article. While they
may have some knowledge about the article's topic (perhaps from the
title), they may not have complete knowledge of its contents, including
the persons or events involved. In this setting, using a low-cost, but
ambiguous referring expression (say a pronoun like ``he'') early may
lead to misunderstanding if context is not informative. But, if by a
later position enough contextual information has accumulated, it may be
efficient to use the ambiguous expression.\par

We consider consider a single language \(\ell\), which contains both
ambiguous and unambiguous utterances. We assume ambiguous utterances are
lower cost. Crucially, we do not assume that the listener knows the
particular topic (\(c_{current}\)) of the conversation
\textit{a priori}. Rather, that the listener has knowledge of the set of
possible topics \(C = \{c_1, \dots, c_k\}\), but does not know which one
is currently being used by the speaker. Formally, this means the
listener does not have access to the correct conditional distribution
over meanings \(P(M|C=c_{current})\) at the start of the discourse.\par

We conduct \(N=600\) simulations, generating discourses of length
\(|D|=30\) utterances, comparing three speaker models (\(n=200\) each).
We consider a single language
\(\ell\)\footnote{See SI for the matrix notation of this langauge.} with
\(|U|=6\) and \(|M|=4\) in which two of the utterances are ambiguous and
lower cost than the unambiguous utterances. (Note that use of this
particular language is not essential -- the results are broadly
generalizable to languages that contain ambiguity, but exploring this
space is computationally expensive.)\par

@@

We consider three types of speaker models. Our \textit{Full pragmatics}
agent, models a speaker who reasons about her listener and also has
complete recall of the set of utterances in the discourse \(D\). This
speaker believes that the listener may not know the current topic
\(c_{current}\) at the start of the discrouse, but can infer it over the
discourse. We compare two baseline models. The first, a
\textit{Partial pragmatics} baseline describes a speaker who reasons
about a listener, but assumes they have no access to discourse history.
The second, a \textit{No pragmatics} speaker who does not consider a
listener at all, but produces utterances according to the underlying
langauge semantics (\(\ell\)) and topic probabilities
(\(p(M|C=c_{current}\)) (see SI section 3).\par

@@

Figure 2 (A) shows the empirical probability that a speaker uses an
ambiguous utterance as a function of discourse position. The
\textit{No pragmatic} baseline uses ambiguous utterances frequently and
at a constant rate over the discourse and the \textit{Partial pragmatic}
baseline avoids ambiguous utterances entirely. But, the
\textit{Full pragmatic} model avoids ambiguous material only at the
start of the discourse, employing it increasingly as the discourse
proceeds. Panel (C) tracks our CE objective for each model over the
discourse. Note that the objective decreases for all three models,
primarily driven by the listener updating his belief about the actual
topic (\(P(C=c_{current}|D)\)). However, the objective declines more
quickly under the \textit{Full} and \textit{Partial} pragmatic speakers
as listener agents are better able to infer the correct context.
Additionally, the difference in CE between the \textit{Full}- and
\textit{Partial pragmatic} models at the end of the discourse is driven
by the reduction in speaker costs (see panel B). While speaker effort
remains constant in both \textit{No pragmatic} and
\textit{Partial pragmatic} baselines, effort declines in the
\textit{Full pragmatic} model as she increasingly relies on ambiguous
material later in the discourse.\par

@@

How do the competing pressures of speakers and listeners give rise to
the distributional regularities found in natural language? Zipf (1949)
proposed that the asymmetry between speaker and listener costs gives
rise to a range of properties at the level of the lexicon. We explored
the interactions of rational pragmatic agents as a framework for
understanding efficient language structure and use. We focused on an
argument on the communicative function of ambiguity (Piantadosi et al.,
2011), deriving a novel speaker-listener Cross-Entropy objective for
measuring the efficiency linguistic systems from first principles of
efficient language use. In a series of simulations we showed that
optimal languages are more likely to contain ambiguous material when
context is informative and that rational pragmatic agents will use
ambiguous material efficiently.\par

A limitation of the current work is an analysis of exactly how this
objective compares to existing measures. For example, previous work has
described competing speaker-listener pressures in terms of a trade-off
of simplicity and informativeness (Kemp \& Regier, 2012) or expressivity
and compressibility (Smith, Tamariz, \& Kirby, 2013). Future work should
assess the degree to which we can derive the same properties as previous
studies using our current framework. In line with this idea, a primary
focus of the current work is to provide a case-study on the productivity
of assessing functionalist theories in repeated reference games between
rational pragmatic agents. We hope this framework can serve as a domain
general tool to assess the range of functionalist theories examining
efficient language-structure and use.\par
